---
title: "NHTS Data Analysis on Drivers and Crashes"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    df_print: paged
  word_document: default
---

#### **Notes**
-   Functions are sourced 
-   NHTS data from 2017. Crashes from 2017 and 2018 - Average of Annual Fatal/Injury Crashes.
-   NHTS: 4.208 million drivers, DMV 4.286 million drivers
-   VMT: NHTS: \~51 (non-commercial travel) WisDOT: 65.324 billion
-   Census: 2017 PEP: Note that 2+ Races - relabeled as Other/Unknown

```{r Setup, include = FALSE}
# Set working directory to where this notebook is saved
knitr::opts_chunk$set(echo = FALSE, comment = NA, warning = FALSE, error = FALSE, message = FALSE, tidy = TRUE, global.par = TRUE, knitr.table.format = "html")
knitr::opts_knit$set(root.dir = '/...')
```

```{r Load Libraries}
library(dplyr)        # to subset, filter, summarize data, in tidyverse, replaces plyr
library(ggplot2)      # create charts
library(tidyverse)    # includes tidyr, tibble, and more
library(lubridate)    # for working with dates, need this for data import
library(expss)        # SPSS style package
library(ggtext)       # HTML for charts
library(data.table)   # db files are loaded as a data table, need this package
library(janitor)      # make tabyl, adorn_totals
# library(rlang)        # {{variable}} to make ggplot2 functions
library(wisdotcrashdatabase)
library(kableExtra)   # styling for HTML
library(summarizeNHTS)
library(censusapi)
library(formattable)

source("W:/HSSA/Keep/Jaclyn Ziebert/Studies and Papers/OWI Moberg/nhts data/Functions_for_NHTS_Analysis_byRaceGenderSex.R") # Source of analysis functions
```

```{r read arrest data}
arrests1718 <- rbind(
  read_csv("W:/HSSA/Keep/Jaclyn Ziebert/Studies and Papers/OWI Moberg/owi convictions data/arrest2017_withcnty.csv"),
  read_csv("W:/HSSA/Keep/Jaclyn Ziebert/Studies and Papers/OWI Moberg/owi convictions data/arrest2018_withcnty.csv"))

arrests1718 <- relabel_arrest(arrests1718)
```

```{r get census data from api}
# census_summarise_pop_count # Census Bureau's Population Estimates Program (PEP) - https://api.census.gov/data/2015/pep/charage https://www.census.gov/programs-surveys/popest/technical-documentation/file-layouts.html
# https://api.census.gov/data/2017/pep/charage/variables.html   # 2017 POP: 5,772,917

# Remove these fields that sum by characteristic
census17 <- getCensus(name = "pep/charage", key = "d1b8a9ed12a716ebe87a794a42b806f5783abf68", vars = c("AGE", "SEX", "RACE", "HISP", "POP"), region = "state:55", vintage = 2017) %>% filter(AGE != "999", SEX != "0",  HISP != "0" , !(RACE %in% c("0", "7", "8", "9", "10", "11")))

census17 = relabel_census(census17)
census17 = census17 %>% get_age_groups_10yr(., age_column = census17$R_AGE)

census17_age18 <- census17 %>% filter(R_AGE >= 18)
```

```{r Run import person crash data and relabel}
persons_columns = c("CRSHNMBR", "CNTYCODE", "CRSHDATE", "TOTUNIT", "UNITNMBR", "CRSHSVR", "WISINJ", "DRVRFLAG", "ROLE", "SEX", "AGE", "RACE", "ALCSUSP", "DRUGSUSP", "DRVRPC", "STATNM")

persons_new <- import_db_data(filepath = "C:/CSV/csv_from_sas/fst/", db_type = "person", columns = persons_columns, years = c("17", "18"))
persons_new <- persons_new %>%
  filter(CRSHSVR != "Property Damage")

persons_new <- relabel_crash_persons(persons_new)
persons_drivers <- persons_new %>% filter(ROLE == "Driver")
```

```{r read raw NHTS data}
# These csvs are to do our own analysis w/o using package functions
nhts_person <- read_csv("C:/NHTS/csv/2017/person.csv") %>% as.data.table() %>% mutate(HOUSEID = as.character(HOUSEID))

# Then we need to relabel variables
nhts_person <- nhts_person %>% nhts_relabel_race() %>% nhts_relabel_race_if_hispanic() %>% nhts_relabel_sex() %>% nhts_relabel_hispanic() %>% nhts_combine_unknown_race() %>% get_age_groups_10yr(., nhts_person$R_AGE) # %>% mutate(R_RACE = factor(R_RACE, levels = race_factor_level), age_group_10yr = factor(age_group_10yr, levels = age_group_10yr_factor_level))

```

```{r read dmv driver}
dmv_drivers <- read.csv("W:/HSSA/Keep/Jaclyn Ziebert/Studies and Papers/OWI Moberg/licensed_drivers/dmv_drivers_1718.csv") %>% as.data.table()
dmv_drivers <- dmv_drivers %>% get_age_groups_10yr(.,dmv_drivers$R_AGE)
```

-   SEX == Unknown and AGE_GROUP == Unknown was removed, so totals among tables may not be equal
-   Ratios are derived by 'share of' of each group

```{r Make TABLE race and age comparisons}
# NOW WITH DMV DRIVERS
# census_df = census17_age18 for adult population  
persons_new %>%
  filter(ROLE == 'Driver') %>%
  get_counts_of_PMD_crash_drivers(., use_dmv_drivers = TRUE, fields_to_summarise = c("R_RACE")) %>%
  arrange(factor(R_RACE, levels  = race_factor_level)) %>% get_data_of_percents_and_ratios_for_table(., fields_that_were_summarised = c("R_RACE")) %>%
  get_column_colors() %>% 
  make_summarized_data_table(name_of_fields_that_were_summarised = c("Race"), caption = "Table 1. Share of travel, driver, and crashes by race")

persons_new %>%
  filter(ROLE == 'Driver') %>% get_counts_of_PMD_crash_drivers(.,  use_dmv_drivers = TRUE, fields_to_summarise = c("age_group_10yr")) %>%
  filter(age_group_10yr != "Unknown") %>% 
  arrange(factor(age_group_10yr, levels  = age_group_10yr_factor_level)) %>%
  get_data_of_percents_and_ratios_for_table(., fields_that_were_summarised = c("age_group_10yr")) %>%
  get_column_colors() %>%
  make_summarized_data_table(name_of_fields_that_were_summarised = c("Age group"), caption = "Table 2. Share of travel, driver, and crashes by age group")

# %>% filter(R_SEX != "Unknown") 
persons_new %>%
  filter(ROLE == 'Driver') %>%
  get_counts_of_PMD_crash_drivers(., use_dmv_drivers = TRUE, fields_to_summarise = c("R_SEX")) %>%
  get_data_of_percents_and_ratios_for_table(., fields_that_were_summarised = c("R_SEX")) %>%
  get_column_colors() %>%
  make_summarized_data_table(name_of_fields_that_were_summarised = c("Sex"), caption = "Table 3. Share of travel, driver, and crashes by sex")

persons_new %>%
  filter(ROLE == 'Driver') %>%
  get_counts_of_PMD_crash_drivers(., use_dmv_drivers = TRUE, fields_to_summarise = c("R_SEX", "R_RACE")) %>%
  filter(R_SEX != "Unknown") %>% 
  arrange(factor(R_RACE, levels  = race_factor_level)) %>%
  get_data_of_percents_and_ratios_for_table(., fields_that_were_summarised = c("R_SEX", "R_RACE")) %>%
  get_column_colors() %>%
  make_summarized_data_table(name_of_fields_that_were_summarised = c("Sex", "Race"), caption = "Table 4. Share of travel, driver, and crashes by sex and race")

persons_new %>%
  filter(ROLE == 'Driver') %>%
  get_counts_of_PMD_crash_drivers(., use_dmv_drivers = TRUE, fields_to_summarise = c("age_group_10yr", "R_RACE")) %>%
  filter(age_group_10yr != "Unknown") %>% 
  arrange(factor(R_RACE, levels  = race_factor_level)) %>%
  get_data_of_percents_and_ratios_for_table(., fields_that_were_summarised = c("age_group_10yr", "R_RACE")) %>%
  get_column_colors() %>% make_summarized_data_table(name_of_fields_that_were_summarised = c("Age group", "Race"), caption = "Table 5. Share of travel, driver, and crashes by age group and race")

persons_new %>%
  filter(ROLE == 'Driver') %>%
  get_counts_of_PMD_crash_drivers(., use_dmv_drivers = TRUE, fields_to_summarise = c("age_group_10yr", "R_SEX")) %>%
  filter(R_SEX != "Unknown", age_group_10yr != "Unknown") %>%
  get_data_of_percents_and_ratios_for_table(., fields_that_were_summarised = c("age_group_10yr", "R_SEX")) %>%
  get_column_colors() %>%
  make_summarized_data_table(name_of_fields_that_were_summarised = c("Age group", "Sex"), caption = "Table 6. Share of travel, driver, and crashes by age group and sex")
```

# Charts

```{r Ratio comparison of race and sex version 2, fig.height=10, fig.width=10}
race_axis_labels <- c('American\nIndian or\nAlaska Native', 'Asian or\nPacific\nIslander', 'Black or\nAfrican\nAmerican', 'Hispanic', 'White', 'Other/\nUnknown') # Took out 'Native Hawaiian or other Pacific Islander'
# By race and sex
ratio_fill = c("#428BCA", "#4DB848", "#D50032","#d6a60f")
persons_new %>% filter(ROLE == 'Driver') %>% get_counts_of_PMD_crash_drivers(., fields_to_summarise = c("R_RACE", "R_SEX")) %>%
  filter(R_SEX != "Unknown") %>%
  get_data_of_percents_and_ratios_for_table(., fields_that_were_summarised = c("R_RACE", "R_SEX")) %>%
  select(R_RACE, R_SEX, c(ends_with("ratio"))) %>%
  # arrange(factor(R_RACE, levels  = race_factor_level)) %>%
  pivot_longer(cols = c(ends_with("ratio")),
               values_to = c("ratio_data")) %>%
  mutate(ratio_data = as.numeric(ratio_data)) %>% 
  filter(R_RACE != "Total") %>%
  ggplot(aes(
    x = factor(R_RACE, levels = race_factor_level),
    y = ratio_data,
    fill = factor(name, levels = ratios_factor_levels)
  )) +
  geom_bar(position = "dodge", stat = "identity") +
  theme_bar_charts(base_size = 12) +
  theme(axis.text.y = element_blank(),
        legend.position =  c(.5, .95),
        legend.direction = "horizontal",
        legend.text=element_markdown(size=12)
        # legend.text = element_text(color =  c( "#428BCA", "#4DB848", "#D50032","#d6a60f"))
  ) +
  # coord_flip() 
  geom_text(
    aes(label = round(ratio_data,1), color = factor(name, levels = ratios_factor_levels)),
    position = position_dodge(width = .9),
    vjust = -.8,
    size = 4,
    fontface = "bold"
  ) +
  scale_y_continuous(expand = expansion(mult = c(0, .05)),
                         name = ""
                         # limits = c(0, 10)
                     ) +
  guides(colour=FALSE) +
  scale_color_manual( # For geom_text
    values = c(
      "crash_drivers_ratio" = "#428BCA",
      "crash_vmt_ratio" = "#4DB848",
      "arrest_driver_ratio" = "#D50032",
      "arrest_vmt_ratio" = "#d6a60f"
    )
  ) + #for geom_text
  scale_fill_manual(
    # For bar chart fill
    # for bars
    name = "",
    labels = paste(
      "<span style='color:",
      ratio_fill,
      ";'>",
      c("Crash to driver", "Crash to VMT", "Arrest to driver", "Arrest to VMT"),
      "</span>"
    ),
    values =  ratio_fill
  ) +
  scale_x_discrete(labels= race_axis_labels) +
  labs(
    title = "Figure 1. Ratios by race and sex"
  ) + facet_wrap( ~ R_SEX, ncol = 1)

```
```{r Ratio comparison of race and age, fig.height=30, fig.width=10}
race_axis_labels <- c('American\nIndian or\nAlaska Native', 'Asian or\nPacific\nIslander', 'Black or\nAfrican\nAmerican', 'Hispanic', 'White', 'Other/\nUnknown') # Took out 'Native Hawaiian or other Pacific Islander'
# By race and sex
persons_new %>% filter(ROLE == 'Driver') %>% get_counts_of_PMD_crash_drivers(., fields_to_summarise = c("R_RACE", "age_group_10yr")) %>%
  filter(age_group_10yr != "Unknown", age_group_10yr != "<16") %>%
  get_data_of_percents_and_ratios_for_table(., fields_that_were_summarised = c("R_RACE", "age_group_10yr")) %>%
  select(R_RACE, age_group_10yr, c(ends_with("ratio"))) %>%
  # arrange(factor(R_RACE, levels  = race_factor_level)) %>%
  pivot_longer(cols = c(ends_with("ratio")),
               values_to = c("ratio_data")) %>%
  mutate(ratio_data = as.numeric(ratio_data)) %>% 
  filter(R_RACE != "Total") %>%
  ggplot(aes(
    x = age_group_10yr,
    y = ratio_data,
    fill = factor(name, levels = ratios_factor_levels)
  )) +
  geom_bar(position = "dodge", stat = "identity") +
  theme_bar_charts(base_size = 12) +
  theme(axis.text.y = element_blank()) +
  # coord_flip() 
  geom_text(
    aes(label = round(ratio_data,1), color = factor(name, levels = ratios_factor_levels)),
    position = position_dodge(width = .9),
    vjust = -.8,
    size = 4,
    fontface = "bold"
  ) +
  scale_y_continuous(expand = expansion(mult = c(0, .05)),
                         name = ""
                         # limits = c(0, 10)
                     ) +
  scale_color_manual( # For geom_text
    values = c(
      "crash_drivers_ratio" = "#428BCA",
      "crash_vmt_ratio" = "#4DB848",
      "arrest_driver_ratio" = "#D50032",
      "arrest_vmt_ratio" = "#d6a60f"
    )
  ) + #for geom_text
  scale_fill_manual( # For bar chart fill
    # for bars
    name = "",
    values = c(
      "crash_drivers_ratio" = "#428BCA",
      "crash_vmt_ratio" = "#4DB848",
      "arrest_driver_ratio" = "#D50032",
      "arrest_vmt_ratio" = "#d6a60f"
    )
    # labels = c("Crashes", "Drivers", "VMT", "Arrest to VMT")
  ) +
  # scale_x_discrete(labels= race_axis_labels) +
  labs(
    title = "<span style='font-size:15pt'>Ratio of
    <span style='color:#428BCA;'>**crash to drivers**</span>,
    <span style='color:#4DB848;'>**crash to VMT**</span>,
    <span style='color:#D50032;'>**arrest to driver**</span>, and
    <span style='color:#d6a60f;'>**arrest to VMT**</span> by race and sex
    </span>"
  ) + facet_wrap(~ factor(R_RACE, levels = race_factor_level), ncol = 1)
```
```{r Ratio comparison of age and sex, fig.height=10, fig.width=10}
# By race and sex
persons_new %>% filter(ROLE == 'Driver') %>% get_counts_of_PMD_crash_drivers(., fields_to_summarise = c("age_group_10yr", "R_SEX")) %>%
  filter(R_SEX != "Unknown", age_group_10yr != "Unknown") %>%
  get_data_of_percents_and_ratios_for_table(., fields_that_were_summarised = c("age_group_10yr", "R_SEX")) %>%
  select(age_group_10yr, R_SEX, c(ends_with("ratio"))) %>%
  # arrange(factor(R_RACE, levels  = race_factor_level)) %>%
  pivot_longer(cols = c(ends_with("ratio")),
               values_to = c("ratio_data")) %>%
  filter(age_group_10yr != "Total") %>%
  mutate(ratio_data = as.numeric(ratio_data)) %>% 
  ggplot(aes(
    x = age_group_10yr,
    y = ratio_data,
    fill = factor(name, levels = ratios_factor_levels)
  )) +
  geom_bar(position = "dodge", stat = "identity") +
  theme_bar_charts(base_size = 12) +
  theme(axis.text.y = element_blank()) +
  # coord_flip() 
  geom_text(
    aes(label = ratio_data, color = factor(name, levels = ratios_factor_levels)),
    position = position_dodge(width = .9),
    vjust = -.8,
    size = 4,
    fontface = "bold"
  ) +
  scale_y_continuous(expand = expansion(mult = c(0, .05)),
                         name = ""
                         # limits = c(0, 4.5)
                     ) +
  scale_color_manual( # For geom_text
    values = c(
      "crash_drivers_ratio" = "#428BCA",
      "crash_vmt_ratio" = "#4DB848",
      "arrest_driver_ratio" = "#D50032",
      "arrest_vmt_ratio" = "#d6a60f"
    )
  ) + #for geom_text
  scale_fill_manual( # For bar chart fill
    # for bars
    name = "",
    values = c(
      "crash_drivers_ratio" = "#428BCA",
      "crash_vmt_ratio" = "#4DB848",
      "arrest_driver_ratio" = "#D50032",
      "arrest_vmt_ratio" = "#d6a60f"
    )
    # labels = c("Crashes", "Drivers", "VMT", "Arrest to VMT")
  ) +
  # scale_x_discrete(labels= race_axis_labels) +
  labs(
    title = "<span style='font-size:15pt'>Ratio of
    <span style='color:#428BCA;'>**crash to drivers**</span>,
    <span style='color:#4DB848;'>**crash to VMT**</span>,
    <span style='color:#D50032;'>**arrest to driver**</span>, and
    <span style='color:#d6a60f;'>**arrest to VMT**</span> by age and sex
    </span>"
  ) + facet_wrap( ~ R_SEX, ncol = 1)
```

```{r chart of age of driver and crash count by VMT}
age_craash_vmt <- persons_new %>% filter(ROLE == "Driver") %>% 
get_counts_of_PMD_crash_drivers(., fields_to_summarise = "age_group_10yr") %>% mutate(byvmt = crash_count/(PMD*total_drivers)*1000000)

age_craash_vmt %>% filter(age_group_10yr != "Unknown") %>% 
ggplot(aes(x= age_group_10yr, y = byvmt)) + geom_bar(fill = "#428BCA", position="dodge", stat="identity") +
  # geom_col(fill = "#428BCA") +
  theme_bar_charts() +
  theme(
    axis.text.x = element_text(size = 14)) +
   geom_text(aes(age_group_10yr, byvmt, label = round(byvmt, 2)),
             color = "#428BCA",
     position = position_dodge(width = .9),
     vjust = -.4,
     size = 5,
     fontface = "bold")+
     scale_y_continuous(limits = c(0, 8),
     expand = expansion(mult = c(0, .05)), name = "") +
  labs(title = "Rate of drivers in a crash by 1,000,000 VMT")
```


```{r chart of age of killed driver by VMT}
age_craash_vmt_killed <- persons_new %>% filter(ROLE == "Driver", WISINJ == 'Fatal Injury') %>% 
get_counts_of_PMD_crash_drivers(., fields_to_summarise = "age_group_10yr") %>% filter(age_group_10yr!= "Unknown") %>% mutate(byvmt = crash_count/(PMD*total_drivers)*1000000)

age_craash_vmt_killed %>%  
ggplot(aes(x= age_group_10yr, y = byvmt)) + geom_bar(fill = "#428BCA", position="dodge", stat="identity") +
  # geom_col(fill = "#428BCA") +
  theme_bar_charts() +
  theme(
    axis.text.x = element_text(size = 14)) +
   geom_text(aes(age_group_10yr, byvmt, label = round(byvmt, 3)),
             color = "#428BCA",
     position = position_dodge(width = .9),
     vjust = -.4,
     size = 5,
     fontface = "bold")+
     scale_y_continuous(limits = c(0, 0.028),
     expand = expansion(mult = c(0, .05)), name = "") +
  labs(title = "Rate of killed drivers in a crash by 1,000,000 VMT")
```
